Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-03-05 manuscript/manuscript.tex
small tweaks to results based on Jesse's comments

\begin{abstract}
\noindent  
Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
\skhcomment{I wonder if we should change this second sentence, because I'm not sure how many of the other cases for which there are actually other methods outside of viruses?Maybe it would be better to say: "For highly diverged sequences, such phylogenetic techniques often estimate surprisingly recent divergence times."Then the next sentence (as already written) talks about how for viruses there are other methods that show these are too recent.}
For highly diverged sequences, such phylogenetic techniques often estimate substantially more recent times than other methods.
In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time due to epistasis. 
\end{abstract}

@@

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, \skhcomment{I realized this citation is wrong. For the fact that ExpCM and mutation-selection models can still incorporate omega, we should cite the relevant Rodrigue / Lartillot paper and Bloom (2017).The Haddox et al (2017) should be cited for the fact that omega can still be drawn from a gamma. At that point, we should also cite relevant Rodrigue / Lartillot paper for their approach to do this.} which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
We will denote such models as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
\ref{tab:empirical_data} shows that all the ExpCM's fit the actual data much better than the GY94 models.
The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
As mentioned in the previous section, $\omega$ generally takes on values $<1$ when fit to all sites in the gene~\citep{murrell2015gene}, and our results follow this trend. 
\ref{tab:empirical_data} shows that all models that fit a single $\omega$ fit very small values, reflecting strong purifying selection against amino-acid change in HA. 
However, the ExpCM's with a single $\omega$ always fit $\omega$ values greater than the GY94 model, suggesting that the site-specific stationary state of the ExpCM's captures some of the purifying selection that the GY94 models can represent only via a small omega.
Among the models with $\Gamma\omega$, the GY94 fits all four $\omega$ categories to values $<< 1$, but the ExpCM's all fit one of the four $\omega$ categories to a value $>1$.
This increase in $\omega$ values makes sense given the different interpretation of $\omega$ in each family of models. 
The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification}\skhcomment{rodrigue MBE}, for amino-acid change, which occurs at some sites in HA due to immune selection.

@@

\begin{figure}
\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
\caption{\label{fig:phylobayes}
\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the ExpCM(H1+H3 avg)+$\Gamma\omega$ and the \modelName{MutSel} model implemented in \texttt{phylobayes}. 
(A) The amino-acid preferences inferred by \modelName{MutSel} are more similar to the average of the H1 and H3 preferences then the H1 and H3 individual preferences sets are to each other, indicating that the average preference set is able to capture the ``global" constraint across the tree.
(B) The ExpCM and the \modelName{MutSel} model estimated similar branch lengths. 
However, the branches near the sequences used in the deep mutationally scanning experiments were estimated to be slightly longer by the ExpCM.
To aid direct comparison between the models, we normalized the branch lengths on each tree by the length between an early and a late seasonal human H1 sequence. 
See Materials and Methods for more information. 
\skhcomment{add in which $\beta$ value I used.}
\skhcomment{I am less sold on using colors on the x or y axis. Is it too confusing with the focal sequence color coding in panel B}
\skhcomment{Order in figure legend doesn't make sense. The first data in the figure is about prefs, not branch lengths, so describe that first. Also, either arrange figure to compare MutSel and avg first, or re-arrange legend to describe them in the order they are presented. I don't know if you need to add exact beta value? Just say re-scaled by stringency parameter in Table X. I don't feel strongly about coloring.}
}
\end{figure}

The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
However, other implementations of mutation-selection models allow their stationary states to be inferred directly from the natural sequence data. 
These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states\skhcomment{overfitting problems associated with trying to make}. 
The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as \modelName{MutSel}. 
The \modelName{MutSel} model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
Like ExpCM's, \modelName{MutSel} models still assume a single set of site-specific amino-acid preferences for the entire tree.

>>>>

\begin{abstract}
\noindent  
Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
For highly diverged sequences, such phylogenetic techniques of- ten estimate surprisingly recent divergence times. 
In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time due to epistasis. 
\end{abstract}

@@

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{rodrigue2016detecting, bloom2017identification}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models~\citep{haddox2017mapping, rodrigue2014site}. 
We will denote such models as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
\ref{tab:empirical_data} shows that all the ExpCM's fit the actual data much better than the GY94 models.
The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
As mentioned in the previous section, $\omega$ generally takes on values $<1$ when fit to all sites in the gene~\citep{murrell2015gene}, and our results follow this trend. 
\ref{tab:empirical_data} shows that all models that fit a single $\omega$ fit very small values, reflecting strong purifying selection against amino-acid change in HA. 
However, the ExpCM's with a single $\omega$ always fit $\omega$ values greater than the GY94 model, suggesting that the site-specific stationary state of the ExpCM's captures some of the purifying selection that the GY94 models can represent only via a small omega.
Among the models with $\Gamma\omega$, the GY94 fits all four $\omega$ categories to values $<< 1$, but the ExpCM's all fit one of the four $\omega$ categories to a value $>1$.
This increase in $\omega$ values makes sense given the different interpretation of $\omega$ in each family of models. 
The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, odrigue2016detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.

@@

\begin{figure}
\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
\caption{\label{fig:phylobayes}
\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the ExpCM(H1+H3 avg)+$\Gamma\omega$ and the \modelName{MutSel} model implemented in \texttt{phylobayes}. 
(A) The amino-acid preferences inferred by \modelName{MutSel} are more similar to the average of the H1 and H3 preferences then the H1 and H3 individual preferences sets are to each other, indicating that the average preference set is able to capture the ``global" constraint across the tree.
(B) The ExpCM and the \modelName{MutSel} model estimated similar branch lengths. 
However, the branches near the sequences used in the deep mutationally scanning experiments were estimated to be slightly longer by the ExpCM.
To aid direct comparison between the models, we normalized the branch lengths on each tree by the length between an early and a late seasonal human H1 sequence. 
See Materials and Methods for more information. 
\skhcomment{Order in figure legend doesn't make sense. The first data in the figure is about prefs, not branch lengths, so describe that first. Also, either arrange figure to compare MutSel and avg first, or re-arrange legend to describe them in the order they are presented. I don't know if you need to add exact beta value? Just say re-scaled by stringency parameter in Table X. I don't feel strongly about coloring.}
}
\end{figure}

The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
However, other implementations of mutation-selection models allow their stationary states to be inferred directly from the natural sequence data. 
These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states.
The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as \modelName{MutSel}. 
The \modelName{MutSel} model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
Like ExpCM's, \modelName{MutSel} models still assume a single set of site-specific amino-acid preferences for the entire tree.
